Anwendungen der internen Sprache von Topoi in der algebraischen Geometrie

### 2017-10-15 slides-phd-defense.tex
Slides for the PhD defense (WIP)

\usepackage[english]{babel}

\usepackage{mathtools}
%\newcommand{\vcentcolon}{:}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
\hypersetup{colorlinks=true}

@@

\title{Using the internal language of toposes in commutative algebra}
\author{Ingo Blechschmidt}
\date{September 17st, 2017}

@@

\newcommand{\pointthis}[2]{%
  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
    (#1) {#1};\node[overlay,rectangle callout,%
    callout relative pointer={(0.5cm,0.5cm)},fill=blue!20] at ($(#1.north)+(-1.1cm,-1.1cm)$) {#2};}%
}%

@@

\newcommand{\slogan}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{2pt}%
    \setlength{\fboxsep}{8pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.93\textwidth}{#1}}}%
  \end{center}%
}

\newcommand{\sloganwithoutborder}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{0pt}%
    \setlength{\fboxsep}{-14pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
  \end{center}%
}

@@

  \small
  ``Without loss of generality, any reduced ring is Noetherian and a field.''

  \hil{Using the internal language of toposes \\ in commutative algebra}
  \medskip

  \scriptsize
  Ingo Blechschmidt \\
  University of Augsburg
  \medskip

  101st Peripatetic Seminar on Sheaves and Logic (Leeds) \\
  September 17st, 2017
  \par
\end{frame}

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
\frame[t]{\frametitle{Quick summary}
  By employing the internal language of toposes in various ways, you can
  pretend that:
  \begin{enumerate}
    \item Sheaves of modules are plain modules.
    \item Schemes are sets: \[ \PP^2_S = \{
    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
    0 \}. \]
    \item Reduced rings are Noetherian and in fact fields.
  \end{enumerate}

% \scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}
}
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

%\frame[t]{\frametitle{Outline}\small
%  \begin{columns}[t]
%    \begin{column}{0.6\textwidth}
%      \vspace*{-1.5em}
%      \begin{itemize}\item[]\tableofcontents\end{itemize}
%      \par
%    \end{column}
%    \begin{column}{0.4\textwidth}
%      \centering
%      \begin{minipage}{0.70\textwidth}
%        \begin{exampleblock}{}
%          \justifying
%          Any finitely generated vector space does \emph{not not} possess a basis.
%        \end{exampleblock}
%      \end{minipage}
%      \medskip
%
%      \scalebox{3}{$\Downarrow$}
%
%      \begin{minipage}{0.70\textwidth}
%        \begin{exampleblock}{}
%          \justifying
%          Any sheaf of modules of finite type on a reduced scheme is locally free
%          \emph{on a dense open subset}.
%        \end{exampleblock}
%        \centering
%        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
%        \par
%      \end{minipage}
%    \end{column}
%  \end{columns}
%}
%
%\note{\justifying\fontsize{8pt}{9.6}\selectfont
%  \begin{center}\large\textbf{Abstract}\end{center}
%
%  \begin{changemargin}{2.5em}{2.5em}
%    (new abstract)
%  \end{changemargin}
%}

\section[Internal language of toposes]{The internal language of toposes}

\subsection{What is a topos?}
\frame[t]{\frametitle{What is a topos?}
  \begin{block}{Formal definition}A \hil{topos} is a category which has finite limits,
  is cartesian closed and has a subobject classifier.
  \end{block}

  \begin{block}{Motto}A topos is a category which is sufficiently rich to support
  an \hil{internal language}.
  \end{block}

  \begin{block}{Examples}\begin{itemize}
    \item \hil{$\Set$:} \tabto{1.4cm} the category of sets
    \item \hil{$\Sh(X)$:} \tabto{1.4cm} the category of set-valued sheaves on a space~$X$
    \item \hil{$\Zar(S)$:} \tabto{1.4cm} the big Zariski topos of a base scheme~$S$
  \end{itemize}\end{block}

  \note{
    \begin{itemize}
      \item \begin{justify}While technically correct, the formal definition is actually
      misleading in a sense: A topos has lots of other vital structure, which
      is crucial for a rounded understanding, but is not listed in the
      definition (which is trimmed for minimality).

      A more comprehensive definition is: A \emph{topos} is a locally cartesian
      closed, finitely complete and cocomplete Heyting category which is exact,
      extensive and has a subobject classifier.\end{justify}

      \item See~\cite{leinster:topos} for a leisurely introduction to topos
      theory.
    \end{itemize}
  }
}

\subsection{Universal localisation}
\frame[t]{\frametitle{Universal localisation}
  \setbeamercolor{block title}{bg=black!30,fg=black}
  \begin{block}{Recall}
    A ring is local iff it has precisely one maximal ideal.

    A homomorphism is local iff it reflects invertibility.
  \end{block}

  Let~$A$ be a ring.
  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.
  \hil{Yes,} if we allow a change of topos:
  Then $A \to A^\sim$ is the universal localisation.
}

\subsection[Kripke--Joyal semantics]{Kripke--Joyal semantics}
\frame[t]{\frametitle{The internal language of a topos}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\E$ \\
      \midrule
      object of~$\E$ & set \\
      morphism in~$\E$ & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      group object & group \\
      \bottomrule
    \end{tabular}
  \end{center}
}

\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
  \small
  Let~$X$ be a topological space. Then we recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  % Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
  \pause
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ and $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{for all open~$V \subseteq U$: } 
    \text{$V \models \varphi$ implies $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
    $U_i \models \varphi(f_i)$}
  \end{array} \]
}

\note{\fontsize{8pt}{9.6}\selectfont
  \begin{itemize}
    \item Special case: The language of~$\Set$ is the usual mathematical language.

    \item \begin{justify}Actually, the objects of~$\E$ feel more like \emph{types} instead of \emph{sets}:
    For instance, there is no global membership relation~$\in$. Rather,
    for each object~$A$ of~$\E$, there is a relation~${\in_A} : A \times \P(A) \to
    \Omega$, where~$\P(A)$ is the power object of~$A$ and~$\Omega$ is the
    object of truth values of~$\E$ (can be understood as the power object of
    a terminal object).\end{justify}

    \item \begin{justify}Compare with the embedding theorem for abelian categories:
    There, an explicit embedding into a category of modules is constructed.
    Here, we only change perspective and talk about the same objects and
    morphisms.\end{justify}

    \item \begin{justify}There exists a weaker variant of the internal language which works
    in abelian categories. By using it, one can even pretend that the objects
    are abelian groups (instead of modules), and when constructing morphisms
    by appealing to the axiom of unique choice (which is a theorem), one
    doesn't even have to check linearity. The proof that this approach
    works uses only categorical logic.\end{justify}

    \item \begin{justify}For expositions of the internal language,
    see Chapters~D1 to~D4 of the Elephant, Chapter~VI of Moerdijk and Mac
    Lane's book, or Chapter~13 of
    \href{http://www.mathematik.tu-darmstadt.de/~streicher/CTCL.pdf}{these
    lecture notes by Thomas Streicher}.\end{justify}
  \end{itemize}
}

\note{\begin{itemize}\justifying
  \item The internal language of a sheaf topos of a $\mathrm{T}_1$-space is
  \emph{classical} (that is, verifies the principle of excluded middle) if and
  only if the space is discrete. That's a not particularly interesting special
  case.
  \item See Section~2.4 of
  \href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes} for remarks on how to appreciate intuitionistic logic.
\end{itemize}}

\note{
  \begin{itemize}
    \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
    formulated over any topos (not just sheaf toposes). They are not all
    arbitrary: Rather, they are very finely concerted to make the crucial
    properties about the internal language (see next slide) true.\end{justify}

    \item \begin{justify}If~$\F$ is an object of~$\Sh(X)$, we write
    ``$f:\F$'' instead of~``$f\in\F$'' to remind us that~$\F$ is not really
    (externally) a set consisting of elements, but that we only pretend this
    by using the internal language.\end{justify}

    \item There are two further rules concerning the constants~$\top$
    and~$\bot$ (truth resp. falsehood):
      \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
        U \models \top &\Ll& U = U \text{ (always fulfiled)}\\
        U \models \bot &\Ll& U = \emptyset
      \end{array} \]

    \item Negation is defined as
    $\neg\varphi :\equiv (\varphi \Rightarrow \bot)$.

    \item \begin{justify}The alternate definition ``$U \models \varphi \vee \psi :\Leftrightarrow
    \text{$U \models \varphi$ or $U \models \psi$}$'' would not be local (cf.
    next slide).\end{justify}
  \end{itemize}
}

\note{
  \begin{itemize}
    \item Let~$\alpha : \F \to \G$ be a morphism of sheaves on~$X$. Then:
    \begin{align*}
      & X \models \speak{$\alpha$ is injective} \\[0.5em]
      \Longleftrightarrow\
      & X \models \forall s,t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
      \Longleftrightarrow\ &
        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
      &\qquad\qquad
          U \models \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
      \Longleftrightarrow\ &
        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
      &\qquad\qquad
          \text{for all open~$V \subseteq U$:} \\
      &\qquad\qquad\qquad\qquad
            \text{$\alpha_V(s|_V) = \alpha_V(t|_V)$ implies $s|_V = t|_V$} \\[0.5em]
      \Longleftrightarrow\ &
        \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
      &\qquad\qquad
            \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
      \Longleftrightarrow\ &
        \text{$\alpha$ is a monomorphism of sheaves}
    \end{align*}

    \item The corner quotes ``$\speak{\ldots}$'' indicate that
    translation into formal language is left to the reader.
  \end{itemize}
}

\note{
  \begin{itemize}
    \item Similarly, we have (exercise, use the rules!):
    \begin{align*}
      &
        X \models \speak{$\alpha$ is surjective} \\[0.5em]
      \Longleftrightarrow\ &
        X \models \forall s\?\G\_ \exists t\?\F\_ \alpha(t) = s \\[0.5em]
      \Longleftrightarrow\ &
        \text{$\alpha$ is an epimorphism of sheaves}
    \end{align*}
  \end{itemize}
}

\note{\fontsize{8pt}{9.6}\selectfont
  \begin{itemize}
    \item One can simplify the rules for often-occuring special cases:
    \[ \renewcommand{\arraystretch}{1.3}\begin{array}{@{}lcl@{}}
      U \models \forall s\?\F\_ \forall t\?\G\_ \varphi(s,t)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$,} \\
      && \text{sections~$s \in \F(V)$, $t \in \G(V)$:} \\
      &&\qquad\qquad V \models \varphi(s,t) \\\\
      U \models \forall s\?\F\_ \varphi(s) \Rightarrow \psi(s)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$, sections~$s \in \F(V)$:} \\
      &&\qquad\qquad \text{$V \models \varphi(s)$ implies $V \models \psi(s)$} \\\\
      U \models \exists!s\?\F\_ \varphi(s)
      &\Longleftrightarrow&
      \text{for all open~$V \subseteq U$,} \\
      &&
      \text{there is exactly one section~$s \in \F(V)$ with:} \\
      &&\qquad\qquad V \models \varphi(s)
    \end{array} \]
  \end{itemize}
}

\note{
  \begin{itemize}
    \item \begin{justify}One can extend the language to allow for \emph{unbounded}
    quantification ($\forall A$ vs.~$\forall a \in A$), by Shulman's
    \emph{stack semantics}. This is needed to formulate
    universal properties internal to~$\Sh(X)$, for instance.\end{justify}

    \item \begin{justify}One can further extend the language to be able to talk
    about locally internal categories over~$\Sh(X)$ (in the sense of
    Penon, see for instance the appendix
    of Johnstone's first topos theory book): Then one can do category theory internal
    to~$\Sh(X)$ using the internal language.

    This specific approach is, as far as I am aware, original work. But of
    course, internal category theory has been done for a long time, see for
    instance the Elephant and also Chapman and Rowbottom's \emph{Relative
    Category Theory and Geometric Morphisms: A Logical Approach}.\end{justify}
  \end{itemize}
}

\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
  \begin{block}{Locality}
    If~$U = \bigcup_i U_i$, then
      $U \models \varphi$ iff $U_i \models
      \varphi$ for all~$i$.
  \end{block}

  \begin{block}{Soundness}
    If~$U \models \varphi$ and if $\varphi$
    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no axiom of choice
    }},
    then~$U \models \psi$.
  \end{block}

@@

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{2.9cm} iff $f|_U = 0 \in \F(U)$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{2.9cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

\note{%
  \vspace{-0.5em}%
  \begin{center}\large\bf Why is constructive mathematics
  interesting?\end{center}
  \vspace{-1em}
  \begin{itemize}
    \item The internal logic of most toposes is constructive.

    \item \begin{justify}From a constructive proof of a statement, it's always possible to
    mechanically extract an \emph{algorithm} witnessing its truth. For example:
    A proof of the infinitude of primes gives rise to an algorithm
    which actually computes infinitely many primes (outputting one at a
    time, never stopping).\end{justify}

    \item \begin{justify}By the celebrated \emph{Curry--Howard correspondence},
    constructive truth of a formula is equivalent to the existence of a
    program of a certain type associated to the formula.\end{justify}

    \item \begin{justify}In constructive mathematics, one can experiment with (and draw
    useful conclusions also holding in a usual sense) \emph{anti-classical
    dream axioms}, for instance the one of synthetic differential geometry:\end{justify}
    \begin{quote}All functions~$\RR \to \RR$ are smooth.\end{quote}

    \item \begin{justify}Constructive accounts of classical theories are sometimes more
    elegant or point out some minor but interesting points which are not appreciated by a
    classical perspective.\end{justify}

    \item \begin{justify}The philosophical question on the \emph{meaning of truth} is
    easier to tackle in constructive mathematics.\end{justify}
  \end{itemize}
}

\note{%
  \vspace{-0.5em}%
  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
  \vspace{-1em}

  \begin{enumerate}
    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
    the term \emph{contradiction} is generally forbidden. This is not the
    case, one has to distinguish between
    \begin{itemize}
      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
    \end{itemize}
    which constructively is only a proof of the weaker
    statement~$\neg\neg\varphi$, and
    \begin{itemize}
      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
      \ldots, contradiction. So~$\neg\psi$ holds.''
    \end{itemize}
    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}

    \item \begin{justify}There is a similar rumour that constructive mathematicians \emph{deny}
    the law of excluded middle. In fact, one can constructively prove that there is no
    counterexample to the law: For any formula~$\varphi$, it holds
    that~$\neg\neg(\varphi \vee \neg\varphi)$.

    In constructive mathematics, one merely doesn't \emph{use} the law of
    excluded middle. (Only in concrete models, for example as provided by the
    internal universe of the sheaf topos on a non-discrete topological space,
    the law of excluded middle will actually be refutable.)\end{justify}
  \end{enumerate}
}

\note{
  \begin{enumerate}
    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
    that most of mathematics breaks down in a constructive setting. This is
    only true if interpreted naively: Often, already very small changes to the
    definitions and statements (which are classically simply equivalent
    reformulations) suffice to make them constructively acceptable.

    In other cases, adding an additional hypotheses, which is classically
    always satisfied, is necessary (and interesting). Here is an example: In
    constructive mathematics, one can not show that any inhabited subset of the
    natural numbers possesses a minimal element. [One can also not show the
    negation -- recall the previous false rumour.] But one can show (quite
    easily, by induction) that any inhabited and \emph{detachable} subset
    of the natural numbers possesses a minimal element. A subset~$U \subseteq
    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
    \in U$ or~$n \not\in U$.

    This has a computational interpretation: Given an arbitrary inhabited subset~$U
    \subseteq \NN$, one cannot algorithmically find its minimal element. But it
    \emph{is} possible if one has an algorithmic \emph{test of membership}
    for~$U$.\end{justify}
  \end{enumerate}

  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
  for references about constructive mathematics. The blog~\cite{bauer:blog} is
  also very informative.\end{justify}
}

\begin{frame}[c]\frametitle{Praise for Mike Shulman}
  \centering
  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
  \par
\end{frame}

\section[The little Zariski topos]{The little Zariski topos of a ring}

\begin{frame}{The little Zariski topos}
  Let~$A$ be a ring. Its \hil{spectrum~$\boldsymbol{\Spec(A)}$} is
  \begin{itemize}
    \item generated by opens~$D(f)$ for~$f \in A$
    \item subject to~$\Spec(A) = \bigcup_i D(f_i)$ iff~$1 = \sum_i g_i f_i$
    for some~$g_i$.
  \end{itemize}

  The \hil{little Zariski topos} of~$A$ is the category~$\Sh(\Spec(A))$ of
  set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
  with~$A^\sim(D(f)) = A[f^{-1}]$, called the \hil{structure sheaf}.

  \begin{block}{Motto (to be amended)}The structure sheaf~$A^\sim$ is a reification of
  all of the stalks~$A_\ppp$.\end{block}

  For instance, all stalks~$A_\ppp$ are integral domains if and only if \[ \Spec(A) \models
  \speak{$A^\sim$ is an integral domain}. \]
\end{frame}

\subsection{Transfer principles}
\begin{frame}{Transfer principles}
  \begin{block}{Theorem}
    The structure sheaf~$A^\sim$ inherits all first-order properties
    of~$A$ which are stable under localisation.
  \end{block}

>>>>

\usepackage[ngerman]{babel}

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
\hypersetup{colorlinks=true}

@@

\title{Anwendungen der internen Sprache von Topoi in der algebraischen
Geometrie}
\author{Ingo Blechschmidt}
\date{October 16th, 2017}

@@

\newcommand{\pointthis}[2]{%
  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
    (#1) {#1};\node[overlay,rectangle callout,%
    callout relative pointer={(-0.2cm,0.5cm)},fill=blue!20] at ($(#1.north)+(0.4cm,-1.1cm)$) {#2};}%
}%

@@

\newcommand{\slogan}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{2pt}%
    \setlength{\fboxsep}{8pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.95\textwidth}{#1}}}%
  \end{center}%
}

\newcommand{\sloganwithoutborder}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{0pt}%
    \setlength{\fboxsep}{-14pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
    text}\parbox{1.02\textwidth}{\begin{center}#1\end{center}}}}%
  \end{center}%
}

@@

  \hil{Anwendungen der internen Sprache von Topoi \\ in der algebraischen
  Geometrie}
  \medskip

  \scriptsize
  Ingo Blechschmidt
  \medskip

  Dissertationsprüfung am \\
  16. Oktober 2017
  \par
\end{frame}

\begin{frame}[plain,c]
  \centering
  \includegraphics[scale=0.37]{images/math-overflow-steven-gubkin}
  \par
\end{frame}

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

\section[Interne Sprache]{Die interne Sprache von Topoi}

\subsection{Was ist ein Topos?}
\frame[t]{\frametitle{Was ist ein Topos?}
  \begin{block}{Formale Definition}Ein \hil{Topos} ist eine kartesisch
  abgeschlossene Kategorie, die alle endliche Limiten und einen
  Unterobjektklassifizierer besitzt.
  \end{block}

  \begin{block}{Motto}Ein Topos ist eine Kategorie, die hinreichend reich ist,
  um eine \hil{interne Sprache} zu erlauben.
  \end{block}

  \begin{block}{Beispiele}\begin{itemize}
    \item \hil{$\boldsymbol{\Set}$:} \tabto{1.5cm} die Kategorie der Mengen
    \item \hil{$\boldsymbol{\Sh(X)}$:} \tabto{1.5cm} die Kategorie der Garben auf einem Raum~$X$
    \item \hil{$\boldsymbol{\Zar(S)}$:} \tabto{1.5cm} der große Zariskitopos eines Basisschemas~$S$
  \end{itemize}\end{block}
}

\subsection{Was ist die interne Sprache?}

\frame[t]{\frametitle{Was ist die interne Sprache?}
  Die interne Sprache eines Topos~$\E$ erlaubt es, in einer \hil{naiven
  Element-basierten Sprache}
  \begin{enumerate}
    \item Objekte und Morphismen des Topos zu konstruieren,
    \item Aussagen über diese zu formulieren und
    \item solche Aussagen zu beweisen.
  \end{enumerate}

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      extern & intern in $\E$ \\
      \midrule
      Objekt von~$\E$ & Menge \\
      Morphismus in~$\E$ & Abbildung zwischen Mengen \\
      Monomorphismus & injektive Abbildung \\
      Epimorphismus & surjektive Abbildung \\
      Gruppenobjekt & Gruppe \\
      \bottomrule
    \end{tabular}
  \end{center}
}
% Interne Sprache von Set nicht vergessen zu erwähnen!

\begin{frame}{Die interne Sprache von~$\boldsymbol{\Sh(X)}$}
  \small
  Sei~$X$ ein topologischer Raum. Dann definieren wir rekursiv
  \[ U \models \varphi \quad \text{("`$\varphi$ gilt auf~$U$"')} \]
  für offene Teilmengen~$U \subseteq X$ und Formeln~$\varphi$.
  \pause
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ und $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ oder $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{es gibt eine Überdeckung $U = \bigcup_i U_i$ sd. für alle~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ oder $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{für alle Offenen~$V \subseteq U$: } 
    \text{$V \models \varphi$ impliziert $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{für alle Schnitte~$f \in \F(V)$ auf~$V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{es gibt eine Überdeckung~$U = \bigcup_i U_i$ sd. für alle~$i$} \\
    && \quad\quad \text{$f_i \in \F(U_i)$ mit
    $U_i \models \varphi(f_i)$ existiert}
  \end{array} \]
\end{frame}

\begin{frame}{Die interne Sprache von~$\boldsymbol{\Sh(X)}$}
  \begin{block}{Lokalität}
    Falls~$U = \bigcup_i U_i$, dann $U \models \varphi$ genau dann, falls $U_i
    \models \varphi$ für alle~$i$.
  \end{block}

  \begin{block}{Korrektheit}
    Falls~$U \models \varphi$ und falls $\varphi$
    \only<1>{konstruktiv}\only<2->{\pointthis{konstruktiv}{
      kein $\varphi \vee \neg\varphi$,\ \
      kein $\neg\neg\varphi \Rightarrow \varphi$,\ \
      kein Auswahlaxiom
    }} $\psi$ impliziert,
    dann~$U \models \psi$.
  \end{block}

@@

  \pause\pause
  \begin{block}{Ein erster Eindruck der konstruktiven Natur}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{2.9cm} genau dann, falls $f|_U = 0 \in \F(U)$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{2.9cm} genau dann, falls~$f = 0$ auf einer dichten offenen
      Teilmenge von~$U$.
  \end{itemize}
  \end{block}
\end{frame}

\section[Kleiner Zariski]{Der kleine Zariskitopos eines Schemas}

\begin{frame}{Der kleine Zariskitopos}
  \begin{block}{Definition}Der \hil{kleine Zariskitopos} eines Schemas~$X$ ist
  die Kategorie~$\Sh(X)$ der mengenwertigen Garben auf~$X$.\end{block}
  \bigskip

  Aus interner Sicht von~$\Sh(X)$ \ldots
  \begin{itemize}
    \item \ \\[-1.2em]\mbox{sieht die Strukturgarbe~$\O_X$ wie ein \hil{gewöhnlicher Ring} und}
    \item \ \\[-1.2em]\mbox{eine Garbe von~$\O_X$-Moduln wie ein \hil{gewöhnlicher Modul}} über
    diesem Ring aus.
  \end{itemize}
\end{frame}

\subsection[Wörterbuch]{Entwicklung eines Wörterbuchs}

\begin{frame}{Entwicklung eines Wörterbuchs}
  \sloganwithoutborder{\hil{Verstehe Konzepte aus der algebraischen Geometrie
  als Konzepte aus der Algebra in der internen Welt von~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \scalebox{0.93}{\begin{tabular}{ll}
      \toprule
      extern & intern in $\Sh(X)$ \\
      \midrule
      Garbe von Mengen & Mengen \\
      Morphismus von Garben & Abbildung zwischen Mengen \\
      Monomorphismus & injektive Abbildung \\
      Epimorphismus & surjektive Abbildung \\
      \midrule
      Garbe von Ringen & Ring \\
      Garbe von Moduln & Modul \\
      Garbe von endlichem Typ & endlich erzeugter Modul \\
      lokal endlich freie Garbe & endlich freier Modul \\
      % coherent sheaf & coherent module \\
      Tensorprodukt von Garben & Tensorprodukt von Moduln \\
      Garbe der Differentialformen & Modul der Kählerdifferentiale \\
      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
      Dimension von $X$ & Krulldimension von~$\O_X$ \\
      \bottomrule
    \end{tabular}}
  \end{center}

\begin{frame}{Anwendungen des Wörterbuchs}
  \vspace*{-2em}
  \begin{columns}[t]
    \begin{column}{0.51\textwidth}
      \begin{center}
        \begin{exampleblock}{}
          \justifying
          Sei~$0 \to M' \to M \to M'' \to 0$ eine kurze exakte Sequenz von
          Moduln. Sind~$M'$ und~$M''$ endlich erzeugt, so auch~$M$.
        \end{exampleblock}
        \medskip

        \begin{exampleblock}{}
          \justifying
          Sei $0 \to \F' \to \F \to \F'' \to 0$ eine kurze exakte Sequenz
          von~$\O_X$-Modulgarben. Sind~$\F'$ und~$\F''$ von endlichem Typ,
          so auch~$\F$.
        \end{exampleblock}
      \end{center}
    \end{column}
    \pause

    \begin{column}{0.39\textwidth}
      \begin{center}
        \begin{exampleblock}{}
          \justifying
          Jeder endlich erzeugte Vektorraum besitzt \emph{nicht nicht} eine
          Basis.
        \end{exampleblock}
        \medskip

        \begin{exampleblock}{}
          \justifying
          Jede Modulgarbe von endlichem Typ auf einem reduzierten Schema ist
          \emph{auf einer dichten offenen Teilmenge} lokal frei.
        \end{exampleblock}
        \centering
        \tiny Ravi Vakil: "`Important hard exercise"' (13.7.K).
        \par
      \end{center}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Mehr zum kleinen Zariskitopos}
  \vspace*{-2em}
  \slogan{\justifying Verstehe Konzepte und Sätze der \hil{algebraischen
  Geometrie} als Konzepte und Sätze der (konstruktiven) \hil{kommutativen
  Algebra} in der internen Welt des \hil{kleinen Zariskitopos}.}

  \begin{itemize}
    \item Transferprinzipien~$M \leftrightarrow M^\sim$
    \item Besondere Eigenschaften der internen Welt
    \item Interne Charakterisierung von Quasikohärenz
    \item Logische Deutung von Ausbreitungsphänomenen
    \item Interne Konstruktion des relativen Spektrums
    \item Interner Beweis von Grothendiecks Lemma über generische Freiheit
  \end{itemize}
\end{frame}

\subsection{Transferprinzipien}

\begin{frame}\frametitle{Transferprinzipien}
  \begin{block}{Theorem}
    Die Strukturgarbe~$\O_{\Spec(A)}$ erbt all solche Eigenschaften erster
    Ordnung von~$A$, die stabil unter Lokalisierung sind.
  \end{block}

  \hil{Beweis.} Die Strukturgarbe~$\O_{\Spec(A)}$ ist die Lokalisierung
  \[ \underline{A}[\F^{-1}] \]
  \justifying
  der konstanten Garbe~$\underline{A}$ am \hil{generischen Filter}~$\F$,
  einer Garbe mit
  $\F_\ppp = A \setminus \ppp$.
  Die Ringe~$A$ und~$\underline{A}$ haben die gleichen Eigenschaften erster
  Ordnung.

  \hil{Bemerkung.} Die analoge Aussage gilt für~$A$-Moduln~$M$ und ihre Spiegelbilder~$M^\sim =
  \underline{M}[\F^{-1}]$.
\end{frame}

\subsection[Besonderheiten]{Besondere Eigenschaften der internen Welt}

\begin{frame}{Besonderheiten der internen Welt}
  Sei~$X$ ein Schema. Intern in~$\Sh(X)$ gilt:
  \begin{center}
    \hil{Jedes nicht-invertierbare Element \\ von~$\boldsymbol{\O_X}$ ist nilpotent.}
  \end{center}
